Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2023-06-29 C - Tutorial Normal Model Rstan/Online Supplement C - Bayesian Hierarchical Modeling in rstan.Rmd
Updates appendices

1. Required packages 
2. Input  
  2.1 Stan file   
  2.2 Data 
3. Fit model 
4. Output   
  4.1 General Effects   
  4.2 Individual Effects 
5. Model comparison 
6. Additional resources

@@

|    As the name says, the Data section specifies the variables we will provide with data. It contains the number of total observations, the number of groups (in our case the number of individuals), the response times, the condition for every response time (side of the digit and the digit indicator) and the prior settings. There are different types of variables. A vector indicates that the variable consists of several numbers and between `[]` the length of the vector is given. An integer indicates that the variable is one full number (i.e., no half numbers such as 0.3, 1.5, 6.33). Real means that the number can be a half number, so it allows numbers as 0.5, 1.3, 5.66, etc. With `<lower>` and `<upper>` the lower and upper bound of the parameter can be specified. For instance, for a variance it could be indicated that this variable cannot be below 0 with `<lower = 0>`.     
|    In the Parameters section, all the parameters that need to be estimated are given. These are the general and individual effects. The specification of the parameters works in the same way as the variable specification in the Data section. Therefore, the general effects are reals, while the individual parameters are vectors with a length of the number of groups (in our case individuals, this means that there will be an estimate for every individual).  
|    The last section, Model, contains the formula to obtain the estimates. We use the target specification as it saves the constants, necessary for model comparison later on. For example, `target += normal_lpdf(delta2 | mu4, sqrt(g4))` basically means that "delta2" is normally distributed with a mean of "mu4" and a standard deviation of "g4". Note that we want to estimate the variances, not the standard deviations. However, in the normal distribution function, the standard deviation is asked for. Therefore, we computed the square root of the variance parameters. Furthermore, it is important that you end every line with `;`, and end the stan file with an empty line. With `//` you can add a comment to the code.

@@

### Individual Effect Correlation

\textcolor{blue}{Up until now, we discussed the normal model from the manuscript. This model assumes that there is no correlation between the individual effects. As stated in the manuscript, it is also possible to allow for the estimation of the correlation between individual effects. This means, for instance, that we would expect that the estimates for the individual effects $\delta_{7,i}$ and $\delta_{3,i}$ are related. To allow for correlation between individual effects, we would have to adjust the formula in the `stan` file and the priors. For the model with correlation terms, we use the file `myModel_cor.stan` that can be found in the folder} *K - R objects > rstan > normal model > correlation*. \textcolor{Notice that we apply a different conceptualization of the model using matrix notation. The effects are called `beta` and the individual effects are called `beta_p`. Because we are using matrices, the effects are stored a bit differently than for the other model. For example, `beta[3]` represents the general digit effect $\delta_{7}$, while `beta_p[1,3]` represents the digit effect for participant 1, $\delta_{7,1}$.}

\textcolor{blue}{When including correlation between individual effects, we are estimating more parameters and have to specify priors for these parameters as well. In specific, we set a prior on the correlation matrices. These priors will be transformed for efficiency and set on so called Cholesky factors of the matrices by the following line of code `Lcorr ~ lkj_corr_cholesky(1)` [for more information on Cholesky factors see} @stan2018a; \textcolor{blue}{section 1.13]. }

@@

To estimate the model with the correlation between random effects, note that one would need to adjust the `file` argument to equal and `myModel_cor.stan`.

Back to our model without the correlation term form the manuscript. We have already fitted the model and saved it in the R object `model_fit`. In case you do not want to wait until the model has been fitted but continue immediately with the rest of this tutorial, you can load this R object that contains the model estimates.  
&nbsp;

@@

  #Define additional settings
  scale_x_continuous(breaks=c(1,2,3,4), labels=c(expression(~delta[~7]), expression(~delta[~6]), expression(~delta[~4]), expression(~delta[~3]))) +
  xlab("") + ylab("Value") +
  ggtitle('') +
  theme_classic() + 
  coord_cartesian(xlim = c(0.75, 5), ylim = c(-0.05, 0.12))

>>>>

1. Required packages 
2. Input  
  2.1 Stan file   
  2.2 Data  
  2.3 Individual Effect Correlation 
3. Fit model 
4. Output   
  4.1 General Effects   
  4.2 Individual Effects 
5. Model comparison 
6. Additional resources

@@

|    As the name says, the Data section specifies the variables we will provide with data. It contains the number of total observations, the number of groups (in our case the number of individuals), the response times, the condition for every response time (side of the digit and the digit indicator) and the prior settings. There are different types of variables. A vector indicates that the variable consists of several numbers and between `[]` the length of the vector is given. An integer indicates that the variable is one full number (i.e., no half numbers such as 0.3, 1.5, 6.33). Real means that the number can be a half number, so it allows numbers as 0.5, 1.3, 5.66, etc. With `<lower>` and `<upper>` the lower and upper bound of the parameter can be specified. For instance, for a variance it could be indicated that this variable cannot be below 0 with `<lower = 0>`.     
|    In the Parameters section, all the parameters that need to be estimated are given. These are the general and individual effects. The specification of the parameters works in the same way as the variable specification in the Data section. Therefore, the general effects are reals, while the individual parameters are vectors with a length of the number of groups (in our case individuals, this means that there will be an estimate for every individual).  
|    The last section, Model, contains the formula to obtain the estimates. We use the target specification as it saves the constants, necessary for model comparison later on. For example, `target += normal_lpdf(delta2 | mu4, sqrt(g4))` means that "delta2" is normally distributed with a mean of "mu4" and a standard deviation of "g4". Note that we want to estimate the variances, not the standard deviations. However, in the normal distribution function, the standard deviation is asked for. Therefore, we computed the square root of the variance parameters. Furthermore, it is important that you end every line with `;`, and end the stan file with an empty line. With `//` you can add a comment to the code.

@@

### \textcolor{blue}{Individual Effect Correlation}

\textcolor{blue}{Up until now, we discussed the normal model from the manuscript. This model assumes that there is no correlation between the individual effects. As stated in the manuscript, it is also possible to allow for the estimation of the correlation between individual effects. This means, for instance, that we would expect that the estimates for the individual effects} $\delta_{7,i}$ and $\delta_{3,i}$ \textcolor{blue}{are related. To allow for correlation between individual effects, we would have to adjust the formula in the} `stan` \textcolor{blue}{file and the priors. For the model with correlation terms, we use the file} `myModel_cor.stan` \textcolor{blue}{that can be found in the folder} *K - R objects > rstan > normal model > correlation*. \textcolor{blue}{Notice that we apply a different conceptualization of the model using matrix notation. The effects are called} `beta` \textcolor{blue}{and the individual effects are called} `beta_p`. \textcolor{blue}{Because we are using matrices, the effects are stored a bit differently than for the other model. For example,} `beta[3]` \textcolor{blue}{represents the general digit effect} $\delta_{7}$, \textcolor{blue}{while} `beta_p[1,3]` \textcolor{blue}{represents the digit effect for participant 1,} $\delta_{7,1}$.}

\textcolor{blue}{When including correlation between individual effects, we are estimating more parameters and have to specify priors for these parameters as well. In specific, we set a prior on the correlation matrices. These priors will be transformed for efficiency and set on so called Cholesky factors of the matrices by the following line of code} `Lcorr ~ lkj_corr_cholesky(1)` \textcolor{blue}{[for more information on Cholesky factors see} @stan2018a; \textcolor{blue}{section 1.13]. }

@@

\textcolor{blue}{To estimate the model with the correlation between individual effects, note that one would need to adjust the} `file` \textcolor{blue}{argument to equal} `myModel_cor.stan`.

\textcolor{blue}{Back to our model without the correlation term form the manuscript.} We have already fitted the model and saved it in the R object `model_fit`. In case you do not want to wait until the model has been fitted but continue immediately with the rest of this tutorial, you can load this R object that contains the model estimates.  
&nbsp;

@@

  #Define additional settings
  scale_x_continuous(breaks=c(1,2,3,4), labels=c(expression(~delta[~7]), expression(~delta[~6]), expression(~delta[~4]), expression(~delta[~3]))) +
  xlab("") + ylab("Value") +
  ggtitle('') +
  theme_classic() + 
  coord_cartesian(xlim = c(0.75, 5), ylim = c(-0.1, 0.15))

### 2023-06-29 D - Tutorial Normal Model brms/Online Supplement D - Bayesian Hierarchical Modeling in brms.Rmd
Updates appendices

1. Required packages 
2. Input  
  2.1 Formula  
  2.2 Family  
  2.3 Data  
  2.4 Priors 
3. Fit model 
4. Output   
  4.1 General Effects   
  4.2 Individual Effects 
5. Model comparison 
6. Additional resources

@@

### Individual Effect Correlation

Up until now, we discussed the normal model from the manuscript. This model assumes that there is no correlation between the individual effects. As stated in the manuscript, it is also possible to allow for the estimation of the correlation between individual effects. This means, for instance, that we would expect that the estimations for the individual effects $\delta_{7,i}$ and $\delta_{3,i}$ are related. To allow for correlation between individual effects, we would have to adjust the model formula. For the estimation of the digit classification task, we could use the following formula:

@@

The difference is that instead of using the double bar `||`, we are now using a single bar `|` before `ind`, indicating that we allow individual effects to correlate.

This means that we are estimating more parameters and therefore we have to specify priors on these parameters as well. In specific, we can set a prior on the correlation matrix called `class = "cor"`. These priors will be transformed for efficiency and set on so called Cholesky factors of the matrices [for more information see @stan2018a; section 1.13]. In our example, we set a prior on the Cholesky factors directly, by defining `class = "L"`, shown in the last line of the code block:

@@

To fit the model with the correlation between random effects, note that one would need to adjust `formula` and `priorsmodel` to the objects where these are saved in (i.e., `bform_cor` and `priorsmodel_cor`).

Back to our model without the correlation term form the manuscript. We have already estimated this model and saved it in the R object `brmmodel8_new03012022.rds`. In case you do not want to wait until the model has been fitted but continue immediately with the rest of this tutorial, you can load this R object that contains the model estimations.  
&nbsp;

@@

>>>>

1. Required packages 
2. Input  
  2.1 Formula  
  2.2 Family  
  2.3 Data  
  2.4 Priors  
  2.5 Individual Effect Correlation 
3. Fit model 
4. Output   
  4.1 General Effects   
  4.2 Individual Effects 
5. Model comparison 
6. Additional resources

@@

### \textcolor{blue}{Individual Effect Correlation}

\textcolor{blue}{Up until now, we discussed the normal model from the manuscript. This model assumes that there is no correlation between the individual effects. As stated in the manuscript, it is also possible to allow for the estimation of the correlation between individual effects. This means, for instance, that we would expect that the estimations for the individual effects} $\delta_{7,i}$ \textcolor{blue}{and} $\delta_{3,i}$ \textcolor{blue}{are related. To allow for correlation between individual effects, we would have to adjust the model formula. For the estimation of the digit classification task, we could use the following formula:}

@@

\textcolor{blue}{The difference is that instead of using the double bar} `||`\textcolor{blue}{, we are now using a single bar} `|` \textcolor{blue}{before} `ind`\textcolor{blue}{, indicating that we allow individual effects to correlate.}

\textcolor{blue}{This means that we are estimating more parameters and therefore we have to specify priors on these parameters as well. In specific, we can set a prior on the correlation matrix called} `class = "cor"`\textcolor{blue}{. These priors will be transformed for efficiency and set on so called Cholesky factors of the matrices [for more information see} @stan2018a; \textcolor{blue}{section 1.13]. In our example, we set a prior on the Cholesky factors directly, by defining} `class = "L"`\textcolor{blue}{, shown in the last line of the code block:}

@@

\textcolor{blue}{To estimate the model with the correlation between individual effects, note that one would need to adjust} `formula` \textcolor{blue}{and} `priorsmodel` \textcolor{blue}{to the objects where these are saved in (i.e.,} `bform_cor` \textcolor{blue}{and} `priorsmodel_cor`\textcolor{blue}{).}

\textcolor{blue}{Back to our model without the correlation term form the manuscript.} We have already estimated this model and saved it in the R object `brmmodel8_new03012022.rds`. In case you do not want to wait until the model has been fitted but continue immediately with the rest of this tutorial, you can load this R object that contains the model estimations.  
&nbsp;

@@

\textcolor{blue}{We can do the same for the model with correlation between the individual effects (see Figure 7) and see that the estimated effects are somewhat different.}

```{r modeldigitestplotcor, echo=FALSE, fig.cap="Model estimates for digit effect parameters in the normal model with correlation between individual effects. The points represent the mean parameter estimates for each individual. The violin plots on the right side show the variance of the individual parameter estimates.", fig.align = "left", fig.height = 11, fig.width = 1, out.width="75%"}
knitr::include_graphics("Fig12_modelestimatefigure2_onlybrms_cor.png")
```
